---
title: "Statistical modelling"
subtitle: "It's a thing that exists"
author: "<br><br>Danielle Navarro &nbsp;`r anicon::faa('twitter', animate='float', rtext='@djnavarro')`"
date: "<br><br>2018-12-11, Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["kunoichi", "ninjutsu" , "assets/ninpo.css"]
    self_contained: false
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(here)
library(tidyverse)
frames <- read_csv(here("analysis","data","frames_ex2.csv"))
```


class: bg-main1 center middle hide-slide-number

.reveal-text.bg-main2[.pad1[
.font4[Linear models]
]]



<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.column.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
tinyframes <- frames %>%
  group_by(id, age, condition) %>%
  summarise(
    response = mean(response)
    ) %>%
  ungroup()

tinyframes
```
]]]
]]

.column.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
tinyframes <- frames %>%
  group_by(id, age, condition) %>%
  summarise(
    response = mean(response)
    ) %>%
  ungroup()

tinyframes
```
]]]
]]



<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.column.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
tinyframes %>%
  group_by(condition) %>%
  summarise(
    mean_resp = mean(response), 
    sd_resp = sd(response),
    n = n()
  )
```
]]]
]]

.column.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
tinyframes %>%
  group_by(condition) %>%
  summarise(
    mean_resp = mean(response), 
    sd_resp = sd(response),
    n = n()
  )
```
]]]
]]



<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.column.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
tinyframes %>%
  group_by(condition) %>%
  summarise(
    mean_resp = mean(response), 
    sd_resp = sd(response),
    n = n()
  )

tinyframes %>% 
  ggplot(aes(
    x = age, 
    y = response, 
    colour = condition)) + 
  geom_smooth(method = "lm") + 
  geom_point()
```
]]]
]]

.column.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
tinyframes %>%
  group_by(condition) %>%
  summarise(
    mean_resp = mean(response), 
    sd_resp = sd(response),
    n = n()
  )

tinyframes %>% 
  ggplot(aes(
    x = age, 
    y = response, 
    colour = condition)) + 
  geom_smooth(method = "lm") + 
  geom_point()
```
]]]
]]




<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.column.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
t.test(
  formula = response ~ condition, 
  data = tinyframes, 
  var.equal = TRUE
)
```
]]]
]]

.column.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
t.test(
  formula = response ~ condition, 
  data = tinyframes, 
  var.equal = TRUE
)
```
]]]
]]


<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.row.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)
```
]]]
]]

.row.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
```
]]]
]]


<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.row.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)

mod2
```
]]]
]]

.row.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)

mod2
```
]]]
]]



<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.row.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)

anova(mod1, mod2)
```
]]]
]]

.row.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)

anova(mod1, mod2)
```
]]]
]]




<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.row.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)
mod3 <- lm(formula = response ~ condition + age, data = tinyframes)

anova(mod1, mod2, mod3)
```
]]]
]]

.row.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)
mod3 <- lm(formula = response ~ condition + age, data = tinyframes)

anova(mod1, mod2, mod3)
```
]]]
]]




<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.row.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)
mod3 <- lm(formula = response ~ condition + age, data = tinyframes)

AIC(mod1, mod2, mod3)
```
]]]
]]

.row.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)
mod3 <- lm(formula = response ~ condition + age, data = tinyframes)

AIC(mod1, mod2, mod3)
```
]]]
]]


<!-- *********** NEW SLIDE ************** -->
---
class: split-40 bg-main1

.row.bg-main1[.content.vtop.center[

.pull.left[.pad1[.font2[
```{r, eval=FALSE, message = FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)
mod3 <- lm(formula = response ~ condition + age, data = tinyframes)

BIC(mod1, mod2, mod3)
```
]]]
]]

.row.bg-main3[.content.vtop.center[
.pull.left[.pad1[.font2[
```{r, eval=TRUE, message = FALSE, echo=FALSE}
mod1 <- lm(formula = response ~ 1, data = tinyframes) 
mod2 <- lm(formula = response ~ condition, data = tinyframes)
mod3 <- lm(formula = response ~ condition + age, data = tinyframes)

BIC(mod1, mod2, mod3)
```
]]]
]]

<!-- *********** NEW SLIDE ************** -->
---
class: bg-main1 center middle hide-slide-number

.reveal-text.bg-main2[.pad1[
.font4[Linear mixed models]
]]



<!-- DONE -->
---

class: bg-main1 middle center

## thank u


