---
title: "R Bootcamp"
subtitle: "Complex Human Data Summer School"
author: "Danielle Navarro & Amy Perfors"
date: "9 Dececember 2018"
output:
  xaringan::moon_reader:
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<!-- the xaringan package -->
<!-- devtools::install_github("yihui/xaringan") -->

<!-- the leaflet package is on CRAN -->


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Welcome!

```{r out.width='100%', fig.height=5.5, eval=require('leaflet')}
library(leaflet)
leaflet() %>% addTiles() %>% setView(144.9612, -37.7963, zoom = 17)
```


<!-- ************************************************ -->
---
class: inverse, middle, center

# Getting Started
http://compcogscisydney.org/psyr/getting-started.html



<!-- ************************************************ -->
---

# Getting started

- Installed R
- Installed RStudio
- Typed a command:

```{r}
345 * 67
```


<!-- ************************************************ -->
---
class: inverse, middle, center

# Variables
http://compcogscisydney.org/psyr/variables.html



<!-- ************************************************ -->
---
# Variables

- Variables are labels for information
- Created by using `<-` 
- Can be numeric, text, and more

--
```{r}
my_name <- "Danielle"
my_age <- 41
```

--
```{r}
my_name
my_age
```


<!-- ************************************************ -->
---
class: inverse, middle, center

# Scripts, Packages
http://compcogscisydney.org/psyr/scripts.html
http://compcogscisydney.org/psyr/packages.html




<!-- ************************************************ -->
---
class: inverse, middle, center

# The Workspace
http://compcogscisydney.org/psyr/workspaces.html



<!-- ************************************************ -->
---
class: inverse, middle, center

# Vectors
http://compcogscisydney.org/psyr/vectors.html


<!-- ************************************************ -->
---
class: inverse, middle, center

# Loops and Conditionals
http://compcogscisydney.org/psyr/loops.html
http://compcogscisydney.org/psyr/branches.html



<!-- ************************************************ -->
---
class: inverse, middle, center

# Writing Functions
http://compcogscisydney.org/psyr/functions.html




<!-- ************************************************ -->
---
class: inverse, middle, center

# A Programming Exercise: <br>The Rescorla-Wagner Model
http://compcogscisydney.org/psyr/programming.html




<!-- ************************************************ -->
---
class: inverse, middle, center

# More R Data Types
http://compcogscisydney.org/psyr/data-types.html







<!-- ************************************************ -->
---
class: inverse, middle, center

# blah blah blah


<!-- this is all stuff from the Xaringan template -->
---

# Some Tips

An example of using the trailing comment `#<<` to highlight lines:

````markdown
`r ''````{r tidy=FALSE}
library(ggplot2)
ggplot(mtcars) + 
  aes(mpg, disp) + 
  geom_point() +   #<<
  geom_smooth()    #<<
```
````

Output:

```{r tidy=FALSE, eval=FALSE}
library(ggplot2)
ggplot(mtcars) + 
  aes(mpg, disp) + 
  geom_point() +   #<<
  geom_smooth()    #<<
```

---

# Some Tips

When you enable line-highlighting, you can also use the chunk option `highlight.output` to highlight specific lines of the text output from a code chunk. For example, `highlight.output = TRUE` means highlighting all lines, and `highlight.output = c(1, 3)` means highlighting the first and third line.

````md
`r ''````{r, highlight.output=c(1, 3)}
head(iris)
```
````

```{r, highlight.output=c(1, 3), echo=FALSE}
head(iris)
```

Question: what does `highlight.output = c(TRUE, FALSE)` mean? (Hint: think about R's recycling of vectors)

